Predicting Lemur's Lifespan

### 2024-11-25 paper/paper.qmd
predictions added updated and scripts cleaned.

\begin{equation}
\bar{a} = \beta_0 + \beta_1 \times \texttt{Sex} + \beta_2 \times \texttt{Species} + \beta_3 \times \texttt{Genus} + \beta_4 \times \texttt{month\_born} + \epsilon
\end{equation}
where,

- $\bar{a}$ is the expected lifespan of a lemur.
- $\beta_0$ is the intercept term. It represents the expected lifespan of a lemur when all other variables are zero.
- $\beta_1$ is the coefficient for `Sex`. It represents the change in the expected lifespan of a lemur for a one-unit change in `Sex`.
- $\beta_2$ is the coefficient for `Species`. It represents the change in the expected lifespan of a lemur for a one-unit change in `Species`.
- $\beta_3$ is the coefficient for `Genus`. It represents the change in the expected lifespan of a lemur for a one-unit change in `Genus`.
- $\beta_4$ is the coefficient for `month_born`. It represents the change in the expected lifespan of a lemur for a one-unit change in `month_born`.
- $\epsilon$ is the error term.

# Appendix 
\appendix

# Appendix A: Species of Lemurs {#appendix_a}
The following are the species of lemurs that are present in the dataset:

>>>>

\begin{equation}
\bar{a} = \beta_0 + \beta_1 \times \texttt{Sex} + \beta_2 \times \texttt{Species} + \beta_3 \times \texttt{Genus} + \beta_4 \times \texttt{month\_born} + \epsilon
\end{equation}

$$
\begin{aligned}
\bar{a}_\text{wild} & \sim \mbox{Normal}(16, 4)\\
\bar{a}_\text{captive} & \sim \mbox{Normal}(26, 4)\\
\beta_0 & \sim \mbox{Normal}(0, 2.5)\\
\beta_1 & \sim \mbox{Normal}(0, 2.5)\\
\beta_2 & \sim \mbox{Normal}(0, 2.5)\\
\beta_3 & \sim \mbox{Normal}(0, 2.5)\\
\beta_4 & \sim \mbox{Normal}(0, 2.5)
\end{aligned}
$$
where,

- $\bar{a}_\text{wild}$ is the expected lifespan of a lemur in the wild.
- $\bar{a}_\text{captive}$ is the expected lifespan of a lemur in captivity.
- $\beta_0$ is the intercept term. It represents the expected lifespan of a lemur when all other variables are zero.
- $\beta_1$ is the coefficient for `Sex`. It represents the change in the expected lifespan of a lemur for a one-unit change in `Sex`.
- $\beta_2$ is the coefficient for `Species`. It represents the change in the expected lifespan of a lemur for a one-unit change in `Species`.
- $\beta_3$ is the coefficient for `Genus`. It represents the change in the expected lifespan of a lemur for a one-unit change in `Genus`.
- $\beta_4$ is the coefficient for `month_born`. It represents the change in the expected lifespan of a lemur for a one-unit change in `month_born`.
- $\epsilon$ is the error term.

We chose a Gaussian family distribution for the model because the dependent variable `Age` is continuous and normally distributed. From @citenjgov we found that the expected lifespan of a lemur in wild is upto 18 years, and in captivity about 30 years. So, we set the prior for the expected lifespan of a lemur in the wild to be $\mbox{Normal}(16, 4)$ and in captivity to be $\mbox{Normal}(26, 4)$. We set the priors for the coefficients to be $\mbox{Normal}(0, 2.5)$, because we do not have any prior information about the effect of the independent variables on the dependent variable.

Using moderately broad priors can help regularize the model, reducing the risk of overfitting and make us have more stable estimates, especially in scenarios with limited data.

@app_model-dataset covers coefficients, standard errors, convergence checks and other model diagnostics for our model.

# Appendix

\appendix

# Data Cleaning {#app_data_cleaning}
I took the following steps to clean the raw data and prepare it for analysis:

1. Loading the Data: The raw data is loaded from the file data/raw_data/animals.csv.

2. Column Selection: The dataset is reduced to relevant columns: `animal_id`, `taxonomic_code`, `sex`, `birth_date`, `death_date`, `birth_type`, `litter_size`, `mother_species`, and `father_species`.

3. Filtering Rows: Rows where death_date is missing are removed to ensure the data includes only animals with complete lifecycle information.

4. Extracting Genus and Species: The taxonomic_code column is split into two new columns: `genus` (first letter) and `species` (remaining characters). The original taxonomic_code column is removed as it is no longer needed.

5. Age Calculation: A new column, `age`, is calculated based on the difference between `death_date` and `birth_date`, measured in years.Lemurs with an age of less than 1 year are removed to focus on mature animals and ignore infant mortality. This was especially common for captive-born animals because excess breeding can lead to high infant mortality rates.

6. Birth Month Extraction: A new column, month_born, is created to indicate the birth month extracted from the birth_date. We needed this information to analyze seasonal effects on lifespan.

7. Splitting Data by Birth Type:
The data is split into two subsets based on birth_type: one for animals born in captivity (`data_captive`) and another for those born in the wild (`data_wild`).

8. Column Removal for Wild Data: Since information on mother_species, father_species, and litter_size is unavailable for wild-born animals, these columns are removed from data_wild.

9. Handling Missing Values: Both datasets, data_wild and data_captive, are cleaned by dropping rows with missing values in key columns: sex, species, age, month_born, and genus.

10. Final Output: The cleaned datasets are saved as CSV files:
Wild-born animals: `data/analysis_data/wild.csv`
Captive-born animals: `data/analysis_data/captive.csv`

This systematic cleaning ensures the dataset is consistent, reliable, and ready for further analysis.

# Analysis Dataset {#app_analysis_dataset}

Hello

# Model Dataset {#app-model-dataset}

#  Species of Lemurs {#appendix_a}
The following are the species of lemurs that are present in the dataset:
